Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-15 paper/paper.tex
Moved proofs to an appendix

@@

% FIXME LF(i, c) picture from the slides for the earlier DBG example
The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. In the worst case, the transformation increases the size of the graph exponentially. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes into BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges into the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph. For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}:
$$
G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
$$
Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$. Due to the nature of de~Bruijn graphs, this index will not produce any false negatives. There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.

@@

\begin{proof}
Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$. The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.

@@

\begin{proof}
If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.

\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
\end{lemma}

\begin{proof}
Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs. Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.

@@

\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

\begin{proof}
(a) For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$. If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct. In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match. Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node. Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.

@@

We implement query $G.\find(X)$ with backward searching. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G.\find(X)$ returns the lexicographic range $[0, \abs{V}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$. We start by mapping the node range $[sp_{i+1}, ep_{i+1}]$ into range
$$
[sp_{in}, ep_{in}] = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]
$$
of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$). Then we map that range into the corresponding range of outgoing edges
$$
[sp_{out}, ep_{out}] = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1].
$$
The lexicographic range containing nodes $G.\find(X[i, \abs{X}-1])$ is
$$
[sp_{i}, ep_{i}] = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
$$

@@

>>>>

Mention that proofs of lemmas are in the appendix

@@

% FIXME LF(i, c) picture from the slides for the earlier DBG example
The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. In the worst case, the transformation increases the size of the graph exponentially. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length. Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph. For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$. Due to the nature of de~Bruijn graphs, this index will not produce any false negatives. There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.

@@

@@

\begin{lemma}[Pruning]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
\end{lemma}

@@

\begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

We implement query $G.\find(X)$ with backward searching. Let $X \in \patternset$ be a pattern. If $\abs{X} = 0$, query $G.\find(X)$ returns the lexicographic range $[0, \abs{V}-1]$ containing all nodes of the path graph. Now assume that $\abs{X} \ge 1$ and that query $G.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$. We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$. We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G.\find(X[i, \abs{X}-1])$:
\begin{align*}
[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1];\ \textrm{and} \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}

@@

\clearpage
\appendix
\section{Proofs of lemmas}

\begin{proof}[Proof of Lemma~\ref{lemma:pg-fn} (No false negatives)]
Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$. The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.

\begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.

\begin{proof}[Proof of Lemma~\ref{lemma:dbg-prune} (Pruning)]
Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs. Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.

\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
(a) The key $G.\gkey(u')$ of a node $u' \in G'.\gpred(v', c)$ prefix-matches string $c \cdot G.\gkey(v')$. If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct. In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match. Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node. Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
